Decline in U.S Birthrates is Indicative of Socio-economic Changes

### 2022-02-27 Output/paper 2.Rmd
Added many changes to text and code

---
title: "Social and Economic Changes have decreased birth rates for millenial and Gen z Mothers (Work in progress)"
author: "Olaedo Okpareke, Arsh"
thanks: 'Code and data are available at: https://github.com/OlaedoOkpareke/Paper2Repository. DOI link:' 
date: "`r format(Sys.time(), '%d %B %Y')`"
abstract: "PUT SOMETHING HERE"
output:
  bookdown::pdf_document2
toc: FALSE
bibliography: references2.bib

---
A nation's birthrate for a particular year is the total number of live births for a particular year. In order to measure the birth rates and therefore fertility levels for the United States, the National Center for Health Statistics (NCHS) aggregates data on birth rates for women by age, race, and educational status. The total population of the country, provided to  to plan public services such as health care, education, and transportation is provided by the CDC SEER and also breaks down the population by age and race. In this report,we are interested in the trend in birth rates for mothers who would be seen today as millenial and generation Z for two different years 20 years apart, and these breakdown in birth rates by race, as well as the relationship between changes in possible explanatory factors like economic, social and cultural aspects of the United states and cganges in these birth rates.

We obtained the dataset from the American Economic Association (cite website). We loaded in, cleaned and analyzed the data using  `R` [@citeR], `dplyr`, [@citedplyr], `tidyr` [@citetidyr], `janitor` [@citejanitor], `haven` [@citehaven], and `tidyverse` [@citetidyverse] packages. Figures and tables were created with `ggplot2` [@citeggplot], `knitr` [@citeknit], `dplyr` [@citedplyr], `usmap` [@citeusmap] and `kableExtra` [@citekable]. We first created graphs such as lineplots, scatterplots and maps which showed the data: the trend in birth rates over the years, both for different age groups and then different races. We then created tables that showed the impact of our chosen groups on total birth rate change, as well as showed the statistics for possible causes or correlations to the birth rate changes.

Young women are not as eager to have children, as this report shows that birth rates have declined for both age groups across the 20 years, and birth rate break downs according to race has also decreased. This implies that the decline is a general national issue as opposed to a specific demogrpahic. The birth rates for the millenial group and Gen z group have gone through significant declines, as well as birth rates for all races experience significant declines. The decline of birth rates for our target groups have also contributed to a significant amount of the total decline in birth rates. Birth rates for our demographic in some stated also experiences large decreases, while some other states had little to no decrease. The variables for birth ratw decline also increased over the time period, correlating to lower birth rates. According to (cite), these rates of birth decline are similar to declines in other high income, developed countries.

While this census is widely used to forecast population growth, biases and problems in data collection could paint an incorrect view of population and birth rates. this data may also systematically hide certain groups of people and their experiences from the ranks, as well as lead to the ignition of certain harmful political ideologies. It could also propogate preconceived biases against women living in poor areas and racial minorities as being "irresponsible" with their bodeis. The declining birth rates brings to light the problems of increasing cost of living, as well as the opportunity costs of working and childcare for women throughout the country, and helps arguments of economic policies that could further hurt the US populace (cite). However the data and statistics also shows a decline in teenage pregnancy, as well as increases in some economic aspects that benefit women. There are also concerns about biases in measurement, ethical collection and inclusion of all, as well as the future state of population growth in the US and its consequences.

@@

The datasets for this report were gotten from the data section for "The Puzzle of Falling US Birth Rates since the Great Recession" (cite), a paper from the Journal of Economic Perspectives. The data used by the paper was compiled by various sources such as the NCHR,CDC SEER, NBER Natality Database, and the Current Population Survey
Annual Social and Economic Supplement (CPS ASEC), the New York Federal Reserve, and the Integrated Public Use Microdata Series (IPUMS) (cite all). The paper and datasets were last accessed on February 16th, 2022. These sources collect and estimate data via national census from US citizens by county or state over the years 1969 to 2019. The birth rate data was gotten from a sample of 100 percent sample of birth certificates (so the population) in some states, and 50 percent of the population as a sample which was randomly chosen in other states (cite). The national population data from the census counts the entire population of the United States by county, thus response was mandatory by government mandate (cite). The datasets compiled by the paper authors typically include demographic data by year and for each state. The data in mainly numeric, counting the number of births, total population, or women in a particular demographic, age group, or population. Meanwhile, categorical data is maimly in the form of state.

@@

For data cleaning, our intended sample was young mothers from two generaions: millenial mothers, i.e women who would be in the age groups of 15-19 or 20-24 in the year 2000, and generation z mothers, who would be 15-19 or 20-24 in the year 2020. The year chosen was 2000 to 2020, to see the change in birthrates from one generation to the other. Millenials are considered to be from 1977 to 1995, and gen z from 1997 to 2012 (cite). We also considered birth rates for white, black, and hispanic women, both for all ages and our target ages in order to compare them. We thus filtered our datasets based on those criteria, or used the closest criteria when they were not available. 2019 is also used if 2020 is not available.

We included a dataset if they were considered important for the question or revealed important information. We included the race dataset as it gave the main information of the paper in the form of birth rate information for different age groups and races. We also included the numbirth dataset as it had information on birthrates for all the age groups together. We used the `birthhist` dataset as it gave the breakdown on birth rates by both age group and race. We used birth1 as well as it had data for young women age groups birth rates, as well as total population data and demographic data for each race and age combination. birth2 had total population data for each age group and race-age combination. We used rent, student_debt, religion, and childcare as they provided information on the change in rental cost, student debt, importance of religion and rise in child costs, which might have a relationship with the birth rates. 
We created several of our own datasets by modifying and joining variables.  These datasets including `mapdiff`, a table which shows the decline of birth rates in each state from 2000 to 2019 and was used to create the map, `fulltable`, which was used to show how our target demogrpahics contributed to birth rate decline, and explanatory, which was used to display the changes in possible explanatory variables over a period of time. Most final datasets include 51 observations (for each state), 102 observations (for each state by year), or 20 observations (for each year).

We modified many variables for the report. We first modified the `numbirth` dataset to add variables for the number of births for each race and age group combination, as they were unavailable in the original dataset. We did this by adding up all educational statuses for each race and age group combination, as each particpant of borth rates was also recorded their race and age status. The final variables were named `numbirthraceage`, aka. `numbirthwhite15` for births for the white 15-19 population. The `birthhist` dataset was gotten by removing all other variables from the numbirth dataset that were not the newly created ones, and then compiling the values by state and year.

We then modified the `race` dataset to change the columns of 15-19 and 20-24 age to rows within the column of `group`. We also modified the `birthhist` dataset's columns to be rows within the column of `racegroup`, then multiplied by 10000 and divided by the total female population to get the birthrates. These variables were used to get the trend in birthrates for our age demographocs, and trend in birthrates in race for our age demographics.

For the `mapdiff` dataset, we obtained the value of `decline`, which showed the change in birth rates by state, by first creating two datasets that got the change in number of birth and population (from the `birth1` and `birth2` datasets respectively) for the 20-34 age group from 1999-2000 and 2018- 2019 by state. The values for the 15-19 and 20-24 ages groups were unavailable for those datasets, so we needed to use the next best age group. From those two new datasets, we changed the year column so that it only stated 2000 or 2019. We then joined those two datasets together and divided the number of births by the population to get the birth rate. We then created two additional datasets that contained only the values of 2000 and only the values of 2019, and colected the mean of each value. To get the final dataset of `mapdiff`, we joined those two datasets by state, and got the `decline` variable by subtracting the average birthrate of 2019 from that of 2000. In this dataset we usd to 20-24 age group as the target age groups specifically were not available.

For the first table, we got the `table` dataset by grouping `numbirth` dataset by year of more than 2000 and then summing up the entire female population by state, which was stored as `sumbirth`. We then created the `ageracepop` dataset using the same methods but now summing up the different age-race combination populations. We named these variables for each combination `sumraceage`, e.g `sumblack20`. We filtered the `birth` dataset between 2000 and 2020. We then joined these three datasets, along with the previous `birthhist`, to create `fulltable`. From `fulltable` we got the `share_of_pop` dataset with variables that got the percentage of population of a demographic by dividing the `sumraceage` variable by the `sumpop` variable. From `fulltable` we also got `birthrates` which got the relative birthrates for each race-age combination, by multiplying the number of births for each race-age combination by 1000 and them dividing them by the total female population. These variables were named `raceagebrate`, e.g `b20brate`. We then got the `contribution` variable by subtracting 2019 values from the 2000 values and dividing that by the `brate_all` for 2000 and multipliying by 200. We then got `percentpop` by multiplying the `share_of_pop` dataset's variables by 100.

For the second table, we got `avg_grossrent` by grouping the `student_debt` variable by year and getting the mean value of all the states. We got `meanchildcare` and `meandebt` using the same methods from the `childcare` and `studentdebt` datasets respectively. For `meandebt`, we multiplied it by 10,000 as that is how it was shown in the original source. We got the `meanpercentimportant` dataset by adding the values of `very important` and `somewhat important` from the `religion` dataset and dividing it by the previous two values and `not_too_important` and `not_at_all_important`, and then getting the mean of all the states. Note that for these variables we needed to use 2004 as 2000 was not available.

@@

knitr::opts_chunk$set(fig.pos = 'H')
library("kableExtra")
install.packages("usmap")
library("tidyverse")
library("haven")
library("janitor")
library("dplyr")
library("usmap")
library("gridExtra")
```

```{r}
birth = read_csv("fig_1.csv")
numbirth = read_csv("numbirths_2001_2019.csv")
student_debt = read_dta("nyfed-student-debt-per-capita-0419.dta")
rent = read_dta("longdiff-rentcost-0419.dta")
religion = read_dta("long-diff-religion.dta")
childcare = read_dta("child_care_costs_under12_0408_1519.dta")
birth2 = read_dta("age_race_comp_seer.dta")
birth1 = read_dta("nchs_births_pop_1990_2019.dta")
race = read_csv("figs_2a_2b.csv")
policy = read_dta("policyvars01_19.dta")
```

STUFF FOR REPRODUCTION

```{r}
birth %>%
  ggplot(data=., aes(x=year, y=brate_all))+
  geom_line(color="blue")+
  theme_classic()+
  geom_vline(xintercept = 2007, linetype="dashed", color = "black", size=0.3) + 
  ylim(50,80) +
  ylab("Births per 1,000 women age 15–44")+
  xlab("")

```{r}
race %>%
  select(-brate_whitenh, -brate_blacknh, -brate_hisp) %>%
  pivot_longer(cols = c("brate_1519", "brate_2024", "brate_2529", "brate_3034", "brate_3539", "brate_4044"),
               names_to = "group",
               values_to = "value"
               ) %>%
  ggplot(mapping = aes(x = year, y = value, color = group)) +
  geom_line() +
  theme_classic() +
  scale_x_continuous(breaks = seq(1980,2020,by = 5))+
  scale_y_continuous(breaks = seq(0,140,by = 20))+
  labs(x = " ",
       y = " ",
       color = " ") +
  geom_vline(xintercept = 2007, linetype="dashed", color = "black", size=0.3) +
  scale_color_brewer(palette = "Set1", labels = c("Age 15-19", "Age 20-24","Age 25-29","Age 30-34","Age 35-39", "Age 40-44")) +
  theme(legend.position = "bottom")
```

```{r}
race %>%
  filter(year >= 1990)%>%
  select(year, brate_whitenh, brate_blacknh, brate_hisp) %>%
  pivot_longer(cols = c("brate_whitenh", "brate_blacknh", "brate_hisp"),
               names_to = "racegroup",
               values_to = "value"
               ) %>%
  ggplot(mapping = aes(x = year, y = value, color = racegroup)) +
  geom_line() +
  theme_classic() +
  scale_x_continuous(breaks = seq(1990,2020,by = 5))+
  ylim(0,140)+
  labs(x = " ",
       y = " ",
       color = " ") +
  scale_color_brewer(palette = "Set1", labels = c("Black, non-Hispanic", "Hispanic", " White, non-Hispanic")) +
  theme(legend.position = "bottom")
```

STUFF FOR ACTUAL PROJECT

We are interested in the trend of birth rates for woman ages 15-19 and 20-24 over a 20 year period. This shows the decline of children and the increase of the average birth age and perhaps increase in abortion. This has possible and already happened implications for public social and economic policy. show a measure of income inequality in housing, and have possible implications for public policy. (Figure \@ref(fig:figone)) and (Figure \@ref(fig:figtwo)) using `ggplot2` [@citeggplot] show the spread and distribution of birth rates for women in the two age groups, subsectioned by race.

These plot show the range of births in the united states according to year
We see that for the 15 to 19 age group, the data is quasi-normal, with most births for the relevant year being between 60 to 190 thousand. The data seems evenly skewed and unimodal. We see that teenage borths tend to be less popular, as the data seems slightly right skewed. This also means that the mean births will be below the median births, which makes sense as teen births in some areas of the US are far higher than others (cite). We also see that black teenaged mothers aare less likely, as they have the lowest births for most years. This is most likely because they are a much smaller segemnt of the population (13% of the population). However we see that black mothers still do manage to have a moderate amount of births. Hispanic teenaged mothers are more middle grtound, having between 60 to 100 thousand births per year. White teenaged mothers are the most likely, do to them being the most in population. They tend to have most of theuir distribution in the high areas, but none in the low areas.

For the 20-24 age group, there seems to be a slight bimodal distribution, with many births being on two extremes. Due to this it is difficult to determine whether the mean or the median of births will be larger.  We see that there are far more women giving birth at this age group than in the previous age group. We see that young women black mothers are overwhelmingly on the lower end of births, which brings into question why teenage blak mothers had more prevalent births. Hispanic mothers also tend to be at the lower end of the scale, but are more spread out. White births are the largest, taking the higher end of the spectrum, indicating their higher populations, or that white women are more likely to have children at these ages.

We see that brths for black and hispanic mothers at age 20-24 are only double the births of their teenaged counterparts, while white mothers tend to be as much as 5times the number of their teenaged counterparts, Therefore, teenaged motherhood in blakc and hispanic women may be more likelty.

In contrast, the distribution for the total population of women for the past 20 years seems uniformal, with alsmost all years having their own value. This makes sense as the population of a nation either increases or decreases, rarely staying the same. Due to this we can say that the mean and median population are approximately the same. However there are two years which had about the same population, this could be due to declining population as a later year has the same population as an earlier year. From the scatterplot we see that the total female population has only been increasing in recent years, with a seemingly exponential increase. This increase in the population could be due to increased amounts of immigration, but is most likely due to the increased lifespan of the elderly to offset the birthrates. The two dips corespind to the previous histogram of two years with the same value.

```{r}
birth1$numbirthwhite15 = birth1$numbirth_lesshs_whitenh_1519 + birth1$numbirth_educ_miss_whitenh_1519 + birth1$numbirth_gtc_whitenh_1519 + birth1$numbirth_somecoll_whitenh_1519 + birth1$numbirth_hsgrad_whitenh_1519 + birth1$numbirth_ba_whitenh_1519

@@

```{r}
birthhist %>%
  pivot_longer(cols = c("white15", "black15","hisp15"),
               names_to = "group",
               values_to = "value"
               ) %>%
  ggplot(data =., aes(x=value, color = group, fill=group)) +
  geom_histogram()+ 
  theme_minimal()+
  xlab("Numbe of births") +
  ylab("Number of Years")

```
```{r}
birthhist %>%
  pivot_longer(cols = c("white20", "black20","hisp20"),
               names_to = "group",
               values_to = "value"
               ) %>%
  ggplot(data =., aes(x=value, color = group, fill=group)) +
  geom_histogram()+ 
  theme_minimal()+
  xlab("Number of Births") +
  ylab("Number of Years")

@@

```{r}
par(mfrow=c(2,2))
pop_plot = birth2 %>%
  group_by(year) %>%
  filter(year >= 2000) %>%
  summarize(population = sum(pop)) %>%
  ggplot(data =., aes(x=population)) +
  geom_histogram(color="black", fill="white",bins=20)+ 
  theme_minimal()+
  xlab("Total Female Population") +
  ylab("Number of Years")

@@

grid.arrange(pop_plot,pop_plot1, ncol =2)
```

@@

(Figure \@ref(fig:figthree)) shows the trend of births for the two age groups in a 20 year period. Millenials would be between ages 15-24 in 2000, and gen Z between ages 15-24 in 2020. We see that there has been a significant decrease in the birth rates, especialy in the 15-19 (teen pregnancy), which has dropped to half its 2000 levels. The 20-24 age grouped by 20 %. This shows the decrease in popualrity for birth at these ages. There was a baby boom in 2006 (cite), which is the reason for the slight increase in birth rates around that time. However overall birth rates for young women has dropped.  According to () there is little reason to beleive these birth rates will increase any time soon. There are some possilbe explanations, including better sex ed for teenagers which led to a decrease in teen pregnancy, mroe women are being focused on their careers and are postponing births, or the increased acceptance of abortions if they can;t afford a child right now.Thesse declining birth rates could also be the cause of declining population (cite), (talk about immigrants due to delcining in discussion). Polcy to increase families can be made to increase the borthrates, and fixing the economic system.

>>>>

---
title: "Decline in U.S Birthrates is Indicative of Socio-economic Changes"
author: "Olaedo Okpareke, Arsh Lakhanpal, Swarnadeep Chattopadhyay"
thanks: 'Code, data and replication are available at: https://github.com/OlaedoOkpareke/Paper2Repository and https://doi.org/10.48152/ssrp-8bs7-7e32' 
date: "`r format(Sys.time(), '%d %B %Y')`"
abstract: "birth rates are merely a symptom of cultursl changes betweeen two generations and the economic situation. It is important to see that these are caused by everything more expensive now. make sure to cite R in the abstract."
output:
  bookdown::pdf_document2
toc: FALSE
bibliography: references2.bib

---
better economic-industrial society to expensive for children, industrial has low birth rates
adoption of more freedom for women?
abortion? restricting of abortion could lead to higher birth rats, but to the detriment of choice

Anyone born between 1978 and 1996 (ages 23 to 38 in 2019) is considered a Millennial, and anyone born from 1997 onward is part of a new

why is hispanic decline so large
harmful ideologies: white people decreasing birht and population, hispanics increasing births and popualtion., hire more immigrants

implications for teenaged mom is a good thing that the borth rates are decreasing?

Check how teen pregnancy  and female enollment has increased.

Its good black teen birth rate decreased (table 1), how can we increase this decrease?

does birth control reduce usage for young mothers?
we dont expect income ratio to affect as these are young moms anyway.
religious change, specially in bible belt?

possible bias: To what extent is random allocation ethical and fair? Some argue that shortages mean it is reasonable to randomly allocate, but that may depend on how linear the benefits are. It may also be difficult to establish definitions, and the power imbalance between those making these decisions and those being treated should be considered

# Introduction

A nation's birthrate for a particular year is the total number of live births per 1000 people for that year. In order to measure the birth rates and therefore fertility levels, as well as forecast population growth for the United States, the National Center for Health Statistics (NCHS) aggregates data on birth rates for women by age, race, educational status, and other factors. The total population of the country is the total number of citizens and other residents in the country, regardless of legal status. It is used in many types of planning, such as planning and decisions on public services, health care, education, government funding, transportation, and politics. The total population is obtained through a population census, provided by the CDC Surveillance, Epidemiolody, and End results program (CDC SEER). This information also breaks down the population by age, race, and socio-economic factors. In this report,we are interested in the trend in birth rates of young women from 2000 to 2020 as well as the breakdown of these birth rates by the race and age demographic of the mother. We are also interested in the relationship between  changes in economic, social and cultural aspects of the United states and the changes in these birth rates.

We obtained the dataset from the American Economic Association [@citewebsite]. We loaded in, cleaned and analyzed the data using  `R` [@citeR], `dplyr`, [@citedplyr], `tidyr` [@citetidyr], `janitor` [@citejanitor], `haven` [@citehaven], and `tidyverse` [@citetidyverse] packages. Figures and tables were created with `ggplot2` [@citeggplot], `knitr` [@citeknit], `dplyr` [@citedplyr], `usmap` [@citeusmap], `gridExtra` [@citegrid] and `kableExtra` [@citekable]. We first created histograms and scatterplots which showed the most important variables. We then created line plots and a map which showed the trend in birth rates over the years, for different age groups, races, and states. We then created tables which showed the relative contributions of our target demographics on the change in birth rates and the differences in their populations, as well as showed the change in values for possible economic and societal factors that may be correlated to birth rates.

Young women are in no rush to have children, as this report shows that birth rates have declined for the 15-24 age groups over the past 20 years. The report also shows that birth rates have declined for all races analyzed over the same period of time.   This implies that this decline is a general national issue attributed to national factors, as opposed to factors that mainly affect a specific demographic. The birth rates for our target demographics have declined to the point that they make up a rather large contribution to the total decline in US birth rates. Birth rates for our age demographics have experienced large decreases in some states, while those in other states have had a relatively smaller decrease or no decrease at all. This may be due to the fact that some economic or social changes of the country affect some geographic areas more than others. These economic and social factors suspected to have an influence on birth rates also changed over the past 20 years in ways that would correlate with declining birth rates. According to [@citeadsera], these rates of birth decline are similar to declines in other high income, developed countries.

While this census is widely used to forecast population growth and make important decisions on the future of the country, biases and problems in data collection could paint an incorrect view of birth rates and the total population. The declining birth rates also brings to light the problems of increasing cost of living, and the increasingly expensive opportunity costs of working and childcare for women throughout the country [@citebarry]. This is problematic as although positive cultural changes have led to more freedom for women than before, as well as decreased teenage births, this also comes with the fact that many are no longer able to afford kids without a strong financial background.The data does not collect information in certain groups of people, thus removing them and their experiences from the ranks, further propagating their systematic discrimination in society. The data collection also helps propagate preconceived biases against young women living in poor, low education areas as well as women of color as being "irresponsible" with their bodies. This is because the data only collects the statistics but does not provide meaning, causes, and explanations behind these statistics [@Gold2001].  There are also concerns about biases in data collection of data, political impacts of the data as well as the future state of population growth in the US and its consequences.

@@

The datasets for this report were gotten from the data section for "The Puzzle of Falling US Birth Rates since the Great Recession" [@citepaper], a paper from the Journal of Economic Perspectives [@citewebsite]. The data used by the paper was compiled by various sources such as the NCHS [@citenchs] CDC SEER [@citecdcseer], NBER Natality Database, and the Current Population Survey
Annual Social and Economic Supplement (CPS ASEC), the New York Federal Reserve, and the Integrated Public Use Microdata Series (IPUMS) (cite all). The paper and datasets were last accessed on February 16th, 2022. The first two sources collect and estimate data via national census from US citizens and residents by county or state over the years 1968 to 2019. The birth rate data was gotten from a sample of 100 percent of birth certificates (i.e the population) in some states, and a sample of 50 percent of the birth certificate in other states, which was randomly sampled [@citenchs]. The national population data from the census counts the entire population of the United States by county, thus response was mandatory by government mandate [@citecdcseer]. The datasets from the other sources compiled by the paper authors typically include data by year and by state for several demographics. These datasets were gotten from surveys or website data, so non-response was often recorded as 0 or as `NA`. The data is mainly numeric, counting the number of births, total population, or women in a particular economic/social/educational demographic, age group, or ethnicity. Meanwhile, categorical data mainly appeared in the form of state.

@@

For data cleaning, our target sample and main focus was young mothers from two generations: millennial mothers, who were between 15 to 24 years in 2000, and generation Z mothers, who were between 15 to 24 years old in 2020. We observed birthrates from 2000 to 2020, in order to see the effect of relatively short term effects on birthrates between the generations. Millennial are often considered to be born between 1980 to 1995, and generation Z from 1996 to 2012 [@citedimock]. We also considered birth rates for white, black, and hispanic women that were in our target age ranges as we wanted to compare the birth rates between them. We thus filtered our datasets based on those criteria or used the closest criteria when they were not available, i.e. 2019 data is used if 2020 data is not available. We dropped any `NA` values from the dataset if they appeared.

## Data Modification

We included a dataset if they were considered important to answering the question behind birthrates, or revealed important information. We included the `race` dataset as it gave the main data of analysis  in the form of birth rates for different age groups and races. We included the `birth` dataset as it included the total birthrate for all women. We also included the `numbirth` dataset as it had information on birthrates for all the age groups combined. We used the `birth1` dataset as well as it had data on birth rates for younger age groups, as well as total population data and demographic data for each race and age combination. `birth2` had total population data for each age group and race-age combination. We used`policy`, `rent`, `student_debt`, `religion`, and `childcare` as they provided information on the changes in unemployment, rental cost, student debt, importance of religion and rise in child costs throughout the US, all of which are expected to influence birth rates to varying degrees.

We created several of our own datasets by modifying and joining variables.  These datasets include `birthhist`, which was gotten from the `birth1` dataset by adding all the educational status for each age-race combination to get the total birth race for that comnination, and then compiling the values by state and year. We also got the `sumpop` variable by adding the total population for each state in a year.  We then obtained `mapdiff`, a table which shows the decline of birth rates in each state from 2000 to 2019 and was used to create the map, `table1`, which was used to show how our target demographics contributed to birth rate decline, and `explanatory`, which was used to display the changes in possible explanatory variables over a period of time. Most final datasets include either 51 observations (for each state), 102 observations (for each state by year), or 20 observations (for each year).

We modified many variables for the report. We first modified the `birth1` dataset to add variables for the number of births for each race and age group combination, as they were unavailable in the original dataset. We did this by adding up all educational statuses for each race and age group combination, as each person whose birth rates was recorded also had their educational status recorded.  The final variables were named `numbirthraceage`, i.e `numbirthwhite15` for births for the white 15-19 population. 
We then modified the `race` dataset to change the columns of 15-19 and 20-24 age to rows within the column of `group`. We also modified the `birthhist` dataset's columns to be rows within the column of `racegroup`, then multiplied by 10000 and divided by the total female population to get the birthrates. These variables were used to get the trend in birthrates for our age demographics, and trend in birthrates in race for our age demographics.

For the `mapdiff` dataset, we obtained the value of `decline`, which showed the change in birth rates by state, by first creating two datasets that got the change in number of births and population (from the `birth1` and `birth2` datasets respectively) for the 20-34 age group from 1999-2000 and 2018- 2019 by state. The values for the 15-19 and 20-24 ages groups were unavailable for those datasets, so we needed to use the next best age group. From those two new datasets, we changed the `year` column so that it only stated 2000 or 2019. We then joined those two datasets together and divided the number of births by the population to get the birth rate. We then created two additional datasets that contained only the values of 2000 and only the values of 2019, and collected the mean of each value. To get the final dataset of `mapdiff`, we joined those two datasets by state, and got the `decline` variable by subtracting the average birthrate of 2019 from that of 2000.

For the first table, we got the `table` dataset by grouping the `numbirth` dataset by year, beyond 2000, and then summing up the entire female population by state, which was stored as `sumbirth`. We then created the `ageracepop` dataset using the same methods but now summing up the different age-race combination populations. We named these variables for each combination `sumraceage`, e.g `sumblack20`. We filtered the `birth` dataset between 2000 and 2020. We then joined these three datasets, along with the previous `birthhist`, to create `fulltable`. From `fulltable` we got the `share_of_pop` dataset with variables that got the percentage of population of a demographic by dividing the `sumraceage` variable by the `sumpop` variable. From `fulltable` we also got `birthrates` which got the relative birthrates for each race-age combination, by multiplying the number of births for each race-age combination by 1000 and them dividing them by the total female population. These variables were named `raceagebrate`, e.g `b20brate`. We then got the `contribution` variable by subtracting 2019 values from the 2000 values and dividing that by the `brate_all` value for 2000 and multiplying by 200. We then got `percentpop` by multiplying the `share_of_pop` dataset's variables by 100.

For the second table, we got `avg_grossrent` by grouping the `student_debt` variable by year and getting the mean value of all the states. We got `meanchildcare` and `meandebt` using the same methods from the `childcare` and `studentdebt` datasets respectively. For `meandebt`, we multiplied it by 10,000 as that is how it was shown in the original source. We got the `meanpercentimportant` dataset by adding the values of `very important` and `somewhat important` from the `religion` dataset and dividing it by the previous two values plus `not_too_important` and `not_at_all_important`, and then getting the mean of all the states. Note that for these variables we needed to use 2004 as 2000 was not available.

@@

knitr::opts_chunk$set(fig.pos = 'H')
library("kableExtra")
# install.packages("usmap")
library("tidyverse")
library("haven")
library("janitor")
library("dplyr")
library("usmap")
library("gridExtra")
```

```{r,include = FALSE}
birth = read_csv(here::here("Inputs/Data/birth.csv"))
numbirth = read_csv(here::here("Inputs/Data/numbirth.csv"))
student_debt = read_csv(here::here("Inputs/Data/student_debt.csv"))
rent = read_csv(here::here("Inputs/Data/rent.csv"))
religion = read_csv(here::here("Inputs/Data/religion.csv"))
childcare = read_csv(here::here("Inputs/Data/childcare.csv"))
birth2 = read_csv(here::here("Inputs/Data/birth2.csv"))
birth1 = read_csv(here::here("Inputs/Data/birth1.csv"))
race = read_csv(here::here("Inputs/Data/race.csv"))
policy = read_csv(here::here("Inputs/Data/policy.csv"))
```
## View of the Data

We are interested in the trend of birth rates for women ages 15-19 and 20-24 over a 20 year period. This shows the decline in popularity of young motherhood and thus the increase of the average birth age. This has ongoing implications for public social and economic policy.  (Figure \@ref(fig:figone)) and (Figure \@ref(fig:figtwo)) using `ggplot2` [@citeggplot] show the spread and distribution of number of births for women in the two age groups for 20 years, sectioned by race.

We see that for the 15 to 19 age group, the data is quasi-normal, with most births for the relevant years being between 60 to 190 thousand. The data seems evenly skewed and unimodal. We see that teenage births tend to be less popular, as the data seems slightly right skewed. This also means that the mean births will be below the median births, which makes sense as teen births in some areas of the US are far higher than others [@Kearney2012]. We also see that black teenage mothers are less common, as they have the lowest births for most years. This is most likely because they are a much smaller segment of the population than the others. However we see that black mothers still do manage to have a moderate amount of births, indicating they are disproportiantely appearing as teenage mothers. Hispanic teenage mothers are more middle ground, having between 60 to 100 thousand births per year. White teenage mothers are the most common, due to them being the largest in population. They tend to have most of their distribution in the high areas towards the higher numbers.

For the 20-24 age group, there seems to be a slight bimodal distribution, with many births being on two extremes. Due to this it is difficult to determine whether the mean or the median of births will be larger.  We see that there are far more women giving birth at this age group than in the previous age group. We see that black mothers are overwhelmingly on the lower end of births, which brings into question why teenage black mothers had more prevalent births compared to their elder counterparts. Hispanic mothers also tend to be at the lower end of the scale, but are more spread out. White births are the largest, taking the higher end of the spectrum, indicating their higher populations, or that white women are more likely to have children at these ages.

We see that births for black and hispanic mothers at age 20-24 are only double the births of their teenage counterparts, while white mothers tend to be as much as 5 times the number of their teenage counterparts, Therefore, teenage motherhood in black and hispanic women may be more common.

In contrast, in (Figure \@ref(fig:figthree)) the distribution for the total population of women for the past 20 years seems uniformal, with almost all years having their own value. This is because the population of a nation either increases or decreases, rarely staying the same. Due to this we can say that the mean and median population are approximately the same. However there are some years which had about the same population, this could be due to declining population as a later year has the same population as an earlier year. From the scatterplot we see that the total female population has only been increasing in recent years, with a seemingly exponential increase. This increase in the population could be due to increased amounts of immigration and the increased lifespan of the elderly to offset the decreasing birthrates. The two dips in the plot correspond to the previous histogram of years with similar value.

```{r, include=FALSE}
birth1$numbirthwhite15 = birth1$numbirth_lesshs_whitenh_1519 + birth1$numbirth_educ_miss_whitenh_1519 + birth1$numbirth_gtc_whitenh_1519 + birth1$numbirth_somecoll_whitenh_1519 + birth1$numbirth_hsgrad_whitenh_1519 + birth1$numbirth_ba_whitenh_1519

@@

```{r figone, echo = FALSE, fig.cap = "Distribution of Number of Births for the 15 to 19 age group from 2000 to 2020"}
birthhist %>%
  pivot_longer(cols = c("white15", "black15","hisp15"),
               names_to = "Race",
               values_to = "value"
               ) %>%
  ggplot(data =., aes(x=value, color = Race, fill=Race)) +
  geom_histogram()+ 
  theme_minimal()+
  xlab("Number of births") +
  ylab("Number of Years")

```
```{r figtwo, echo = FALSE, fig.cap = "Distribution of the number of births from the 20 to 24 age group from 2000 to 2020" }
birthhist %>%
  pivot_longer(cols = c("white20", "black20","hisp20"),
               names_to = "Race",
               values_to = "value"
               ) %>%
  ggplot(data =., aes(x=value, color = Race, fill=Race)) +
  geom_histogram()+ 
  theme_minimal()+
  xlab("Number of Births") +
  ylab("Number of Years")

@@

```{r figthree, echo = FALSE, fig.cap = "Distribution of the total number of births and total female population from 2000 to 2020"}
par(mfrow=c(2,2))
pop_plot = birth2 %>%
  group_by(year) %>%
  filter(year >= 2000) %>%
  summarize(population = sum(pop)) %>%
  ggplot(data =., aes(x=population)) +
  geom_histogram(color="black", fill="white",bins=20)+ 
  theme_minimal()+
  xlab("Total Female Population") +
  ylab("Number of Years")

@@

grid.arrange(pop_plot,pop_plot1, nrow =2)
```

@@

(Figure \@ref(fig:figfour)) shows the trend of births for the two age groups in a 20 year period. We see that there has been a significant decrease in the birth rates, especially in the 15-19 age group, dropping to half of its 2000 levels. This indicates a decrease in teenage pregnancy nationwide. The birth rates for the 20-24 age dropped by around 40% over the same time period. This may be because generation Z women are deciding to have children later in life. There was a baby boom in 2006 [@citecatholic], which is the reason for the slight increase in birth rates around that time. However overall birth rates for young women has dropped.  According to [@Caldwell2006] there is little reason to believe these birth rates will increase any time soon. There are some possible explanations behind this phenomenon, including better sex education for teenagers leading to a decrease in teen pregnancy, more women focusing on their careers and thus postponing births [@citeexecutive], or the increased acceptance of abortions. These declining birth rates could also be the cause of declining population. There are also economic factors correlated to this decline, including the rising cost of living.

```{r figfour, echo = FALSE, fig.cap = "Trend in Birth rates for the 15-19 and 20-24 age groups from 2000 to 2020"}

race %>%
  select(year,brate_1519,brate_2024) %>%
  filter(year >= 2000) %>%
  pivot_longer(cols = c("brate_1519", "brate_2024"),
               names_to = "group",
               values_to = "value"
               ) %>%
  ggplot(mapping = aes(x = year, y = value, color = group)) +
  geom_line() +
  theme_minimal() +
  scale_x_continuous(breaks = seq(2000,2020,by = 2))+
  scale_y_continuous(breaks = seq(0,140,by = 20))+
  labs(x = "Year",
       y = "Birth Rate",
       color = "Age group") +
  scale_color_brewer(palette = "Set1", labels = c("Age 15-19", "Age 20-24")) +
  theme(legend.position = "bottom")
```

(Figure \@ref(fig:figfive)) shows the trend of births for different races of women from the ages of 15 to 24 over the past 20 years. We see that the birthrate for Hispanic mothers significantly decreased, going from about 63 births per 1000 population to just about 40 births for 1000 population. This can either be because the population of hispanic women has increased due to immigration, or that the number of births has decreased. The birth rates for black women has also halved, going from 50 births per 1000 to just 25 births per 1000. The white birth rates however has taken the largest decrease, going from about 125 births to about 65 births through the 20 years. We see that while millennial mothers (mothers in 2000) had a wide spread in the birth rates for different races, for generation z they have converged more closely due to the decrease in birth rates for all races. This could be due to economic factors directly related to race, such as inuequality in the healthcare system [@Davis2018] . As the income for hispanic families increases and they become more westernized, it is possible that they may be adhering to the western standard of fewer children [@immigbirth].
